Supermartingales and Termination Analysis of Statistical PCF

### 2021-01-23 LICS21/SPCF.tex
Minor reformatting in appendix; added reference

\iffalse
\lo{Alternatively, define the \emph{runtime of $M$} to be the random variable 
\[
T_M(s) := 
\begin{cases}
\min \set{n \mid \pi_0(\red^n(M, s)) \textrm{ is a value}} & \hbox{if $\red^\infty(M,s)$ is defined}\\
\infty & \hbox{otherwise}
\end{cases}
\]
Equivalently, we say that $M$ is \emph{almost surely terminating} (AST) if $T_M < \infty$ a.s.; 
and $M$ is \emph{positively almost surely terminating} (PAST) if $\expect{T_M} < \infty$.}
\fi
For example, the term 
\[
(\tY \lambda f n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)} ) \, \underline{0},
\] 
which generates a geometric distribution, terminates on the set $\entrosp \setminus [0.5,1]^\mathbb N$, which has measure 1, therefore it terminates almost surely, whereas 
\(
\tif{\tsample - 0.5 < 0}{\underline 0}{(\tY \lambda x. x) \, \underline 0}
\), 
which terminates on the set $\pi_h^{-1}[[0,0.5]]$, has probability 0.5 of failing to terminate.

%\paragraph{}
%The requirement that $M$ be closed is not important for the proofs of the theorems, but is simply needed to ensure that $M$ reaches a value, rather than a general normal form like $x \, y$. \lo{This can be omitted.}

>>>>

\iffalse
\lo{Alternatively, define the \emph{runtime of $M$} to be the random variable 
\[
T_M(s) := 
\begin{cases}
\min \set{n \mid \pi_0(\red^n(M, s)) \textrm{ is a value}} & \hbox{if $\red^\infty(M,s)$ is defined}\\
\infty & \hbox{otherwise}
\end{cases}
\]
Equivalently, we say that $M$ is \emph{almost surely terminating} (AST) if $T_M < \infty$ a.s.; 
and $M$ is \emph{positively almost surely terminating} (PAST) if $\expect{T_M} < \infty$.}
\fi
\begin{example} The term 
\[
(\tY \lambda f n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)} ) \, \underline{0},
\] 
which generates a geometric distribution, terminates on the set $\entrosp \setminus [0.5,1]^\mathbb N$, which has measure 1, therefore it terminates almost surely; whereas 
\(
\tif{\tsample - 0.5 < 0}{\underline 0}{(\tY \lambda x. x) \, \underline 0}
\), 
which terminates on the set $\pi_h^{-1}[[0,0.5]]$, has probability 0.5 of failing to terminate.
\end{example}
%\paragraph{}
%The requirement that $M$ be closed is not important for the proofs of the theorems, but is simply needed to ensure that $M$ reaches a value, rather than a general normal form like $x \, y$. \lo{This can be omitted.}

### 2021-01-23 LICS21/antitone-proofs.tex
Minor reformatting in appendix; added reference

\Cref{ex:raven complex}
Let $\Theta := \tY \, \lambda f x . \tif{x \leq 0}{0}{f(x - \tsample)}$. 
We can construct a sparse ranking function $f$ for $\Theta \, \underline{10}$ as follows:
\begin{align*}
\Theta \, \underline l 
&\mapsto 
l + 2
\\
\tif{\underline l \leq 0}{0}{\Theta \, (\underline l - \tsample)}
&\mapsto
l + 1
\\
\underline 0 &\mapsto 0.
\end{align*}

@@

\Cref{ex:Fair-in-the-limit random walk}
\[
\underbrace{\big
(\tY \, \lambda f x .\ 
\tif{x \leq 0}{0}{f(x - 1) \oplus_{\frac{x}{2x+1}} f(x + 1)} \big)}_{\Xi} 
\, \underline{10}
\]
To construct an antitone ranking function, we solve the recurrence relation:
% \[
% \left\{
% \begin{array}{rll}
% z_0 &=& 0\\
% n \geq 1, \quad z_n &>& \dfrac{n}{2n + 1} \, z_{n-1} + \dfrac{n+1}{2n + 1} \, z_{n+1}.
% \end{array}
% \right.
% \]
\begin{align*}
z_0 &= 0\\
n \geq 1, \quad z_n &> \textstyle \frac{n}{2n + 1} \, z_{n-1} + \frac{n+1}{2n + 1} \, z_{n+1}.
\end{align*}
For $n > 2$, $z_n = \ln(n-1)$ works. The expected decrease is $\frac 1 2(\ln(1+\frac 1 {(n-1)^2-1}) - \frac 1 {2n+1} \ln(1 + \frac 2 {n-2}))$, and using the fact that $\frac x {x+1} \leq \ln (1+x) \leq x$ for $x > 0$, this is at least $\frac 1 2(\frac 1 {(n-1)^2} - \frac 1 {2n + 1} \frac 2 {n-2}) = \frac {n^2}{2(n-1)^2(2n+1)(n-2)}$, which (again for $n > 2$) is positive and antitone. For $n = 0, 1, 2$, we take $\epsilon$ to be 9/40 (the same as its value at 3), then set $z_2 = 2 \ln 2 - \ln 3 - 1, z_1 = 3 \ln 2 - 2 \ln 3 - 3, z_0 = 4 \ln 2 - 3 \ln 3 - 6$. Some of those values are negative, but it's still bounded below, so by adding a constant offset it can be corrected, and the term is AST.

@@

\subsubsection*{Non-affine recursion}

@@

\medskip

\Cref{ex:non-affine recursion easy}
Let 
\[
\Xi := \tY \lambda f x. x \oplus_{2/3} f (f (x + 1))
\] 
A suitable sparse ranking function for $\Xi \, \underline 1$ would be $\Xi^n \, \underline i \mapsto 3 \, n$ (for $n \geq 0$), because $\Xi^n \, \underline i$ reduces to either $\Xi^{n+1} \, \underline{i+1}$ or $\Xi^{n-1} \, \underline{i+1}$, with probabilities $1/3$ and $2/3$. The value of $i$ does not actually matter for the progress of this recursion. It is basically another variant of the biased random walk, except that the relevant variable is the number of copies of $\Xi$, instead of a real number in the term.

@@

\subsubsection*{Higher-order recursion}

@@

\Cref{ex:higher-order recursion}
Consider the higher-order function
$\Xi : (\textsf{R} \to \textsf{R} \to \textsf{R}) \to \textsf{R} \to \textsf{R} \to \textsf{R}$
recursively defined by
\[
\begin{array}{l}
\Xi := \tY \lambda \varphi \, f^{\textsf{R} \to \textsf{R} \to \textsf{R}} \, s^\textsf{R} \, n^\textsf{R} . \\
\quad\tif{n \leq 0}{s}{
f \, n \, (\varphi \, f \, s \, (n - 1))
\oplus_p
f \, n \, (\varphi \, f \, s \, (n + 1))
}
\end{array}
\]
Let $F : \textsf{R} \to \textsf{R} \to \textsf{R}$ be a function such that $F\ \underline n\ \underline m$ terminates with no $\tY$ reductions for all $m$ and $n$. Then $\Xi\ F\ \underline x\ \underline y$ has the antitone sparse ranking function $F \underline{n_0} (F \underline{n_1} (\dots(\Xi\ F\ \underline x\ \underline n)\dots) \mapsto g_2(n)$, where $g_2$ is as in \Cref{ex:unbiased random walk}, using the same antitone function too.

>>>>

\subsection*{\Cref{ex:raven complex}}
Let $\Theta := \tY \, \lambda f x . \tif{x \leq 0}{0}{f(x - \tsample)}$. 
We can construct a sparse ranking function $f$ for $\Theta \, \underline{10}$ as follows:
\begin{align*}
\Theta \, \underline l 
&\mapsto 
l + 2
\\
\tif{\underline l \leq 0}{0}{\Theta \, (\underline l - \tsample)}
&\mapsto
l + 1
\\
\underline 0 &\mapsto 0.
\end{align*}

@@

\subsection*{\Cref{ex:Fair-in-the-limit random walk}}
\[
\underbrace{\big
(\tY \, \lambda f x .\ 
\tif{x \leq 0}{0}{f(x - 1) \oplus_{\frac{x}{2x+1}} f(x + 1)} \big)}_{\Xi} 
\, \underline{10}
\]
To construct an antitone ranking function, we solve the recurrence relation:
% \[
% \left\{
% \begin{array}{rll}
% z_0 &=& 0\\
% n \geq 1, \quad z_n &>& \dfrac{n}{2n + 1} \, z_{n-1} + \dfrac{n+1}{2n + 1} \, z_{n+1}.
% \end{array}
% \right.
% \]
\begin{align*}
z_0 &= 0\\
n \geq 1, \quad z_n &> \textstyle \frac{n}{2n + 1} \, z_{n-1} + \frac{n+1}{2n + 1} \, z_{n+1}.
\end{align*}
For $n > 2$, $z_n = \ln(n-1)$ works. The expected decrease is $\frac 1 2(\ln(1+\frac 1 {(n-1)^2-1}) - \frac 1 {2n+1} \ln(1 + \frac 2 {n-2}))$, and using the fact that $\frac x {x+1} \leq \ln (1+x) \leq x$ for $x > 0$, this is at least $\frac 1 2(\frac 1 {(n-1)^2} - \frac 1 {2n + 1} \frac 2 {n-2}) = \frac {n^2}{2(n-1)^2(2n+1)(n-2)}$, which (again for $n > 2$) is positive and antitone. For $n = 0, 1, 2$, we take $\epsilon$ to be 9/40 (the same as its value at 3), then set $z_2 = 2 \ln 2 - \ln 3 - 1, z_1 = 3 \ln 2 - 2 \ln 3 - 3, z_0 = 4 \ln 2 - 3 \ln 3 - 6$. Some of those values are negative, but it's still bounded below, so by adding a constant offset it can be corrected, and the term is AST.

@@

\subsection*{Non-affine recursion}

@@

\subsection*{\Cref{ex:non-affine recursion easy}}
Let 
\[
\Xi := \tY \lambda f x. x \oplus_{2/3} f (f (x + 1))
\] 
A suitable sparse ranking function for $\Xi \, \underline 1$ would be $\Xi^n \, \underline i \mapsto 3 \, n$ (for $n \geq 0$), because $\Xi^n \, \underline i$ reduces to either $\Xi^{n+1} \, \underline{i+1}$ or $\Xi^{n-1} \, \underline{i+1}$, with probabilities $1/3$ and $2/3$. The value of $i$ does not actually matter for the progress of this recursion. It is basically another variant of the biased random walk, except that the relevant variable is the number of copies of $\Xi$, instead of a real number in the term.

@@

\subsection*{Higher-order recursion}

@@

\subsection*{\Cref{ex:higher-order recursion}}
Consider the higher-order function
$\Xi : (\textsf{R} \to \textsf{R} \to \textsf{R}) \to \textsf{R} \to \textsf{R} \to \textsf{R}$
recursively defined by
\[
\begin{array}{l}
\Xi := \tY \lambda \varphi \, f^{\textsf{R} \to \textsf{R} \to \textsf{R}} \, s^\textsf{R} \, n^\textsf{R} . \\
\quad\tif{n \leq 0}{s}{
f \, n \, (\varphi \, f \, s \, (n - 1))
\oplus_p
f \, n \, (\varphi \, f \, s \, (n + 1))
}
\end{array}
\]
Let $F : \textsf{R} \to \textsf{R} \to \textsf{R}$ be a function such that $F\ \underline n\ \underline m$ terminates with no $\tY$ reductions for all $m$ and $n$. Then $\Xi\ F\ \underline x\ \underline y$ has the antitone sparse ranking function $F \underline{n_0} (F \underline{n_1} (\dots(\Xi\ F\ \underline x\ \underline n)\dots) \mapsto g_2(n)$, where $g_2$ is as in \Cref{ex:unbiased random walk}, using the same antitone function too.

### 2021-01-23 LICS21/intro.tex
Minor reformatting in appendix; added reference

The AST problem is not just important, but also difficult: 
Deciding AST of first-order imperative programs with discrete probabilities is $\Pi^0_2$-complete \cite{DBLP:conf/mfcs/KaminskiK15}.
In recent unpublished work (under double-blind review elsewhere), the AST problem for higher-order programs with continuous distributions (and suitable primitive functions) is shown to have the same complexity.
Though the problem of proving AST of imperative probabilistic programs is much studied \cite{DBLP:journals/jacm/KaminskiKMO18,DBLP:conf/lics/OlmedoKKM16,DBLP:conf/mfcs/KaminskiK15,DBLP:series/mcs/McIverM05,DBLP:conf/popl/FioritiH15,DBLP:conf/pldi/ChenH20,DBLP:journals/toplas/ChatterjeeFNH18}, to our knowledge the problem of proving AST of higher-order programs with continuous distributions is new.

>>>>

The AST problem is not just important, but also difficult: 
Deciding AST of first-order imperative programs with discrete probabilities is $\Pi^0_2$-complete \cite{DBLP:conf/mfcs/KaminskiK15}.
In recent work (under double-blind review elsewhere), the AST problem for higher-order programs with continuous distributions (and suitable primitive functions) is shown to have the same complexity.
Though the problem of proving AST of imperative probabilistic programs is much studied \cite{DBLP:journals/jacm/KaminskiKMO18,DBLP:conf/lics/OlmedoKKM16,DBLP:conf/mfcs/KaminskiK15,DBLP:series/mcs/McIverM05,DBLP:conf/popl/FioritiH15,DBLP:conf/pldi/ChenH20,DBLP:journals/toplas/ChatterjeeFNH18}, to our knowledge the problem of proving AST of higher-order programs with continuous distributions is new.

### 2021-01-23 LICS21/supermartingales.tex
Minor reformatting in appendix; added reference

It will be demonstrated later that for any rankable term $M$, if $(M_n)_{n \geq 0}$ is the reduction sequence starting from $M$ (considered as a stochastic process), then $(f(M_n))_{n \geq 0}$ is a supermartingale, and $M$ terminates almost surely, but first, some preliminaries about supermartingales (see e.g.~\cite{Williams1999} for details).
\iffalse
\lo{It is confusing (strictly speaking, incorrect) to say that $(M_n)_{n \geq 0}$ is a reduction sequence from $M$. Formally $M_n$ is the random variable $M_n: s \mapsto \pi_0 (\red^n(M, s))$.}
\akr{I guess it's implicitly identifying a sequence of functions ($(M_n)_{n \geq 0}$), and a function producing sequences (a reduction-sequence-valued random variable), but this sort of thing seems very standard in dealing with random variables (e.g.~applying a function to some random variables, which is really a sort of concatenation).}
\fi
%\paragraph{}

>>>>

It will be demonstrated later that for any rankable term $M$, if $(M_n)_{n \geq 0}$ is the reduction sequence starting from $M$ (considered as a stochastic process), then $(f(M_n))_{n \geq 0}$ is a supermartingale, and $M$ terminates almost surely, but first, some preliminaries about supermartingales (see e.g.~\cite{Williams1999,Durrett2019} for details).
\iffalse
\lo{It is confusing (strictly speaking, incorrect) to say that $(M_n)_{n \geq 0}$ is a reduction sequence from $M$. Formally $M_n$ is the random variable $M_n: s \mapsto \pi_0 (\red^n(M, s))$.}
\akr{I guess it's implicitly identifying a sequence of functions ($(M_n)_{n \geq 0}$), and a function producing sequences (a reduction-sequence-valued random variable), but this sort of thing seems very standard in dealing with random variables (e.g.~applying a function to some random variables, which is really a sort of concatenation).}
\fi
%\paragraph{}
